---
title: "Explore Paris Paintings"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Choose two of the following questions and answer as a team.

- What are some important features of the distribution of the prices of paintings sold?
- Describe the distribution of schools of paintings?
- Are the paintings sold mostly by living artists?
- How do prices vary over time and by school of painting?
- Is a particular type of buyer more represented than other?
- Could we think of any way to categorize the paintings from high- to low-end? 
What could be some criteria that help us construct this categorization?

## Accessing the data

- Codebook: https://stat.duke.edu/courses/Fall15/sta112.01/data/paris_paintings.html

- Go to the [Resources on Sakai](https://sakai.duke.edu/portal/site/a91015e2-86df-49b6-9c16-e83a9ca5da66/page/fb4fe022-ccfe-4806-a4ff-4d73fdb1bf08) and download `paris_paintings.csv`

- Upload this file to RStudio Server

- Load using the following (make sure data file is in the correct working
directory):

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

## Submission instructions

Your submission should be an R Markdown file in your team App Ex repo, in a folder 
called `AppEx_02`.

## Due date

Thursday, Sep 17, beg of class

## Watch out for...

merge conflics on GitHub -- you're working in the same repo now!

Issues will arise, and that's fine! Commit and push often, and ask questions
when stuck.