---
title: "Inference for a proportion"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tasks

### 1. Organ donors

Re-write the code from the organ donors example from class to

- use as few hard coded arguments as possible (the ultimate 
goal is to be able to re-use the code with little or no modification 
for the next dataset/hypothesis test)
- use 15,000 simulations
- plot the null distribution as a histogram
- calculate the p-value in R (as opposed to counting number of simulations off of plot)
- write up the conclusion based on the new p-value (and make sure your interpretation/
conlusion is in context of the data and the research question)
    - use inline R code to report the p-value in the text (if you're not sure how,
    see the "Inline R code" section at http://rmarkdown.rstudio.com)

For reference, here is the organ donors example:

>People providing an organ for donation sometimes seek the help of a special "medical 
>consultant". These consultants assist the patient in all aspects of the surgery, with 
>the goal of reducing the possibility of complications during the medical procedure and 
>recovery. Patients might choose a consultant based in part on the historical 
>complication rate of the consultant's clients. 
>
>One consultant tried to attract patients by noting that the average complication rate 
>for liver donor surgeries in the US is about 10%, but her clients have only had 3 
>complications in the 62 liver donor surgeries she has facilitated. She claims this is 
>strong evidence that her work meaningfully contributes to reducing complications (and 
>therefore she should be hired!).

and here is the code we developed in class:

```{r eval=FALSE}
sim_dist <- data.frame(p_hat_sim = rep(NA, 100))
for (i in 1:100){
  sim <- sample(outcomes, size = 62, prob = c(0.1, 0.9), replace = TRUE)
  p_hat_sim <- sum(sim == "complication") / length(sim)
  sim_dist$p_hat_sim[i] <- p_hat_sim
}

ggplot(sim_dist, aes(x = p_hat_sim)) + 
  geom_dotplot()
```

### 2. Planned Parenthood funding

A September 2015 [Pew Research report](http://www.people-press.org/2015/09/28/majority-says-any-budget-deal-must-include-planned-parenthood-funding/) states

> The latest national poll by the Pew Research Center, conducted Sept. 22-27 among 
> 1,502 adults, finds that 60% say that any budget deal must maintain funding for Planned 
> Parenthood, while 32% say that any agreement must eliminate funding for the organization.

The title of this report is "Majority Says Any Budget Deal Must Include Planned Parenthood 
Funding".

Do these data provide convincing (statistically significant) evidence for this title?

Answer this question using the code you developed in the previous exercises with
as little modification as possible. As you are working on this exercise, you might
find that you need to go back and improve your code to eliminate hardcoded objects.

As usual, report a p-value and state your conclusion in context of the data and the 
research question. Once again, use inline R code for incorporating the p-value in
the text.

## Submission instructions

Your submission should be an R Markdown file in your (new) team App Ex repo, in a folder 
called `AppEx_06`.

## Due date

Tuesday, Oct 20, begginning of class

## Watch out for...

... merge conflics on GitHub -- you're working in the same repo now!

Issues will arise, and that's fine! Commit and push often, and ask questions
when stuck.