---
title: "Selecting models for Paris Paintings"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Task: Fit models that include both main and interaction effects and do model selection. 

#### Breaking it down:

1. Decide on a subset of variables to consider for your analysis. Think 
about it as focusing on certain aspects of the price determination, as opposed to 
all aspects. Talk to the data expert for guidance on which variables pertain to
specific aspects of paintings.
    - You're allowed a maximum of 10 total variables to consider, including
    interactions.

2. Decide among these which variables might make sense to interact. Once again, you'll
need expert guidance on this. Remember, we consider interactions when variables might 
behave differently for various levels of another variable. Some interactions of potential
interest are:

    - School of painting & landscape variables: `school_pntg` & `landsALL` / `lands_figs` /
    `lands_ment`
    - Landscapes & paired paintings: `landsALL` / `lands_figs` / `lands_ment` & `paired`
    - Other artists & paired paintings: `othartist` & `paired`
    - Size & paired paintings: `surface` & `paired`
    - Size & figures: `surface` & `figures` / `nfigures`
    - Dealer & previous owner: `dealer` & `prevcoll`
    - Winning bidder & prevcoll: `endbuyer` & `prevcoll`
    - Winning bidder & artist living: `winningbiddertype` & `artistliving`

This is not an exhaustive list, so you might come up with others.

3. Use backwards elimination to do model selection. Make sure to show 
each step of decision (though you don't have to interpret models at each stage).
    - Provide interpretations for the slopes for the final model you arrive at 
    and create at least one visualization that supports your narrative.

**Bonus:** You can also get creative with new / composite variables. 

#### Weekend milestone: 

Check in with Sandra on Piazza about which variables make sense to consider 
together, interact, etc. to tell a coherent story about the prices of these 
paintings. At least 1 post per team by Sunday evening. Take advantage of this
resource for building meaningful models and interpretations!

## Accessing the data

- Codebook: https://stat.duke.edu/courses/Fall15/sta112.01/data/paris_paintings.html

- Go to the [Resources on Sakai](https://sakai.duke.edu/portal/site/a91015e2-86df-49b6-9c16-e83a9ca5da66/page/fb4fe022-ccfe-4806-a4ff-4d73fdb1bf08) and download `paris_paintings.csv`

- Upload this file to RStudio Server

- Load using the following (make sure data file is in the correct working
directory):

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

## Submission instructions

Your submission should be an R Markdown file in your team App Ex repo, in a folder 
called `AppEx_05`.

## Due date

Thursday, Sep 29, beg of class

## Watch out for...

merge conflics on GitHub -- you're working in the same repo now!

Issues will arise, and that's fine! Commit and push often, and ask questions
when stuck.