---
title: "Modeling prices of Paris Paintings - a simple approach (first)"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task

Your main goal is to model the relationship between various variables in the dataset
and price of paintings. For this application exercise you are asked to stick
to models with one explanatory variable.

- Model fitting:
    - Try at least 5 different models, that is at least 5 different explanatory
variables.
    - Make sure some are categorical and some are numerical.
    - **Extra credit:** For the adventorous... Re-level at least one of your 
    variables, e.g. combine a few levels into one, or turn a numerical variable into a 
    categorical variable. When doing this, do not overwrite the original variable, but 
    instead create a new variable. *Hint:* `plyr` package: `revalue()` or 
    `mapvalues()` or `car` package: `recode()` might be useful, but there are other
    ways of accomplishing this too.

- Interpretations:
    - Interpret the slope and the intercept of each model.
    - Interpret the $R^2$ of each model.

- Synthesis: At the end write one synthesis paragraph comparing your models and 
determine which model does the best job in explaining the variability in prices
of paintings. Your interpretations should be in context of the data, which means
you need to understand the context of your data. Thankfully your data expert will
be available to answer questions on Piazza! (But don't leave them till the last
minute.)

## Accessing the data

- Codebook: https://stat.duke.edu/courses/Fall15/sta112.01/data/paris_paintings.html

- Go to the [Resources on Sakai](https://sakai.duke.edu/portal/site/a91015e2-86df-49b6-9c16-e83a9ca5da66/page/fb4fe022-ccfe-4806-a4ff-4d73fdb1bf08) and download `paris_paintings.csv`

- Upload this file to RStudio Server

- Load using the following (make sure data file is in the correct working
directory):

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

## Submission instructions

Your submission should be an R Markdown file in your team App Ex repo, in a folder 
called `AppEx_03`.

## Due date

Tuesday, Sep 22, beg of class

## Watch out for...

merge conflics on GitHub -- you're working in the same repo now!

Issues will arise, and that's fine! Commit and push often, and ask questions
when stuck.