---
title: "Modeling log transformed prices of Paris Paintings"
output: 
  html_document: 
    highlight: pygments
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task

#### Price vs. some binary variable:
- Fit a model predicting `log(price)` from one of the binary variables
in the dataset.
- Write the linear model, in the form $\widehat{y} = b_0 + b_1 x$ but using
the actual variables instead of $y$ and $x$, and using the estimated coefficients
using $b_0$ and $b_1$.
- Interpret the slope and the intercept.
- Calculate and interpret $R^2$.

#### Price vs. some numerical variable:
- Fit a model predicting `log(price)` from one of the numerical variables
in the dataset.
- Write the linear model, same as above.
- Interpret the slope and the intercept.
- Calculate and interpret $R^2$.

#### Price vs. material:
- Recreate the recoding of the material variable: `mat_recode` (from class)
- Fit a model predicting `log(price)` from `mat_recode`.
- Write the linear model, same as above.
- Interpret the slopes and the intercept.
- Calculate and interpret $R^2$.
- Paintings on which material type are predicted to be the most expensive?

#### Synthesis: 
At the end write one synthesis paragraph comparing your models and 
determine which model does the best job in explaining the variability in prices
of paintings. Your interpretations should be in context of the data, which means
you need to understand the context of your data. Thankfully your data expert will
be available to answer questions on Piazza! (But don't leave them till the last
minute.)

## Tip

Keep interpretations concise!

## Accessing the data

- Codebook: https://stat.duke.edu/courses/Fall15/sta112.01/data/paris_paintings.html

- Go to the [Resources on Sakai](https://sakai.duke.edu/portal/site/a91015e2-86df-49b6-9c16-e83a9ca5da66/page/fb4fe022-ccfe-4806-a4ff-4d73fdb1bf08) and download `paris_paintings.csv`

- Upload this file to RStudio Server

- Load using the following (make sure data file is in the correct working
directory):

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

## Submission instructions

Your submission should be an R Markdown file in your team App Ex repo, in a folder 
called `AppEx_04`.

## Due date

Thursday, Sep 24, begginning of class

## Watch out for...

... merge conflics on GitHub -- you're working in the same repo now!

Issues will arise, and that's fine! Commit and push often, and ask questions
when stuck.