---
title: "Sta112FS <br> 7. The language of models"
author: "Dr. Çetinkaya-Rundel"
date: "September 17, 2015"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
    css: ../lec.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```

# Today's agenda

## Today's agenda

- Review App Ex from last time

- Modeling the relationship between variables
    - Focus on *linear* models (but remember there are other types of models too!)

- Application Exercise: model prices of Paris Paintings

- **Due Tuesday:** Finish App Ex + Reading (you'll receive an email with
a link after class)

# Prepping the data

## Load packages + Paris Paintings data

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(stringr)
```

```{r echo=FALSE}
pp <- read.csv("../../Data/paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```


## What's going on with prices? {.build}

```{r}
class(pp$price)
```

```{r}
table(pp$price)[1:30]
```

## Let's first fix those prices

Replace `,` with `` (blank), and save the variable as numeric:
```{r}
pp <- pp %>%
  mutate(price = as.numeric(str_replace(price, ",", "")))
```

Much better...
```{r}
class(pp$price)
```

## Prices

<div class="question">
Describe the distribution of prices of paintings.
</div>

```{r, fig.height=4, fig.width=7}
ggplot(data = pp, aes(x = price)) +
  geom_histogram(binwidth = 1000)
```

# Modeling the relationship between variables

## Models as functions

- We can represent relationships between variables using **function**

- A function is a mathematical concept: the relationship between an output
and one or more inputs. 
    - Plug in the inputs and receive back the output
    - Example: the formula $y = 3x + 7$ is a function with input $x$ and output $y$,
    when $x$ is $5$, the output $y$ is $22$
    
## Height as a function of width

```{r warning = FALSE, fig.height=4, fig.width=7}
ggplot(data = pp, aes(x = Width_in, y = Height_in)) +
  geom_point() +
  stat_smooth(method = "lm") # lm for linear model
```
    
## Vocabulary

- **Response variable:** Variable whose behavior or variation you are trying to understand, on the y-axis (dependent variable)

- **Explanatory variables:** Other variables that you want to use to explain the variation in the response, on the x-axis (independent variables)

- **Model value:** Output of the function **model function**
    - The model function gives the typical value of the response variable
    *conditioning* on the explanatory variables
    - Also called the **predicted value**
    
- **Residuals:** Show how far each case is from its model value
    - $residual = actual~value - model~value$
    - Tells how far above the model function each case is

## Residuals

<div class="question">
What does a negative residual mean? Which paintings on the plot have have negative 
residuals?
</div>

```{r warning = FALSE, fig.height=4, fig.width=7, echo=FALSE}
ggplot(data = pp, aes(x = Width_in, y = Height_in)) +
  geom_point() +
  stat_smooth(method = "lm")
```

## Multiple explanatory variables

<div class="question">
How, if at all, the relatonship between width and height of paintings vary by whether
or not they have any landscape elements?
</div>

<div class="columns-2">
```{r warning = FALSE, fig.height=4, fig.width=6, echo=FALSE}
ggplot(data = pp, aes(x = Width_in, y = Height_in, color = factor(landsALL))) +
  geom_point(alpha = 0.4) +
  stat_smooth(method = "lm")
```

```{r warning = FALSE, fig.height=4, fig.width=4.5, echo=FALSE}
ggplot(data = pp, aes(x = Width_in, y = Height_in)) +
  geom_point(alpha = 0.4) +
  stat_smooth(method = "lm")
```
</div>

## Just for reference...

Here is the code for the two plots in the previous slide

```{r warning = FALSE, fig.height=4, fig.width=6, eval=FALSE}
# points colored by landsALL type
ggplot(data = pp, aes(x = Width_in, y = Height_in, color = factor(landsALL))) +
  geom_point(alpha = 0.4) +
  stat_smooth(method = "lm")
```

```{r warning = FALSE, fig.height=4, fig.width=4.5, eval=FALSE}
# points not colored by landsALL type
ggplot(data = pp, aes(x = Width_in, y = Height_in)) +
  geom_point(alpha = 0.4) +
  stat_smooth(method = "lm")
```

## Models - upsides and downsides

- Models can sometimes reveal patterns that are not evident in a graph of the
data. This is a great advantage of modeling over simple visual inspection of
data. 

- There is a real risk, however, that a model is imposing structure that is
not really there on the scatter of data, just as people imagine animal shapes in
the stars. A skeptical approach is always warranted.

## Variation around the model...

is just as important as the model, if not more!

*Statistics is the explanation of variation in the context of what remains
unexplained.*

- The scatter suggests that there might be other factors that account for large parts 
of painting-to-painting variability, or perhaps just that randomness plays a big role.

- Adding more explanatory variables to a model can sometimes usefully reduce
the size of the scatter around the model. (We'll talk more about this later.)

# How do we use models?

1. Explanation: Characterize the relationship between $y$ and $x$ via *slopes* for
numerical explanatory variables or *differences* for categorical explanatory
variables

2. Prediction: Plug in $x$, get the predicted $y$

# Characterizing relationships with models

## Relationship between height and width {.smaller}

```{r}
lm(Height_in ~ Width_in, data = pp)
```

\[ \widehat{Height_{in}} = 3.62 + 0.78~Width_{in} \]

- **Slope:** For each additional inch the painting is wider, the height is expected
to be higher, on average, by 0.78 inches.

- **Intercept:** Paintings that are 0 inches wide are expected to be 3.62 inches high,
on average.
    - Does this make sense?
    
## Relationship between height and landscape features {.smaller}

```{r}
lm(Height_in ~ factor(landsALL), data = pp)
```

\[ \widehat{Height_{in}} = 22.68 - 5.65~landsALL \]

- **Slope:** Paintings that have some landscape features are expected, on average,
to be 5.65 inches shorter than paintings that don't have landscape features.
    - Compares the baseline level (`landsALL = 0`) to the other level
    (`landsALL = 1`).

- **Intercept:** Paintings that don't have landscape features are expected, on 
average, to be 22.68 inches tall.

## Relationship between height and school {.smaller}

```{r}
lm(Height_in ~ school_pntg, data = pp)
```

- When the categorical explanatory variable has many levels, they're encoded to
**dummy variables**.

- Each coefficient describes the expected difference between heights in that 
particular school compared to the baseline level.

## Correlation does not imply causation!

Remember this when interpreting model coefficients

# Prediction with models

## Predict height from width {.build}

<div class="question">
On average, how tall are paintings that are 60 inches wide?
\[ \widehat{Height_{in}} = 3.62 + 0.78~Width_{in} \]
</div>

```{r}
3.62 + 0.78 * 60
```

"On average, we expect paintings that are 60 inches wide to be 50.42 inches high."

**Warning:** We "expect" this to happen, but there will be some variability. (We'll
learn about measuring the variability around the prediction later.)

## Prediction vs. extrapolation

<div class="question">
On average, how tall are paintings that are 400 inches wide?
\[ \widehat{Height_{in}} = 3.62 + 0.78~Width_{in} \]
</div>

```{r warning = FALSE, fig.height=4, fig.width=6, echo=FALSE}
ggplot(data = pp, aes(x = Width_in, y = Height_in)) +
  geom_point(alpha = 0.4) +
  stat_smooth(method = "lm", fullrange = TRUE) +
  xlim(0, 400) +
  ylim(0, 320)
```

## Watch out for extrapolation!

"When those blizzards hit the East Coast this winter, it proved to my satisfaction 
that global warming was a fraud. That snow was freezing cold. But in an alarming 
trend, temperatures this spring have risen. Consider this: On February 6th it was 10 
degrees. Today it hit almost 80. At this rate, by August it will be 220 degrees. So 
clearly folks the climate debate rages on." - Stephen Colbert April 6th, 2010

Source: OpenIntro Statistics. "Extrapolation is treacherous." OpenIntro Statistics.

# Measuring model fit

## Measuring the strength of the fit

- The strength of the fit of a linear model is most commonly evaluated using $R^2$

- It tells us what percent of variability in the response variable is explained by 
the model.

- The remainder of the variability is explained by variables not included in the 
model.

- $R^2$ is sometimes called the coefficient of determination.

## Obtaining $R^2$ in R {.build .smaller}

- Height vs. width
```{r}
m_ht_wt <- lm(Height_in ~ Width_in, data = pp) # fit and save
summary(m_ht_wt)$r.squared # extract R-squared
```

Roughly 68% of the variability in heights of paintings can be explained by their
widths.

- Height vs. lanscape features
```{r}
m_ht_land <- lm(Height_in ~ landsALL, data = pp)
summary(m_ht_land)$r.squared
```

## Attribution

Some of the material on these slides is based off of

Daniel Kaplan. Statistical Modeling: A Fresh Approach. (2009).

# Application exercise

## App Ex 3

See course website