---
title: "Sta112FS <br> 8. Recoding variables and transformations"
author: "Dr. Çetinkaya-Rundel"
date: "September 22, 2015"
output:
  html_document:
    highlight: pygments
    css: ../lec.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 90)
```

# Today's agenda

## Today's agenda

- Review App Ex from last time
    - Recap modeling terminology and interpretations

- Recoding variables
    - So you can make them more meaningful and use them in your analysis
    
- Transformations
    - So you actually fit linear models to linear relationships

- **Due Thursday:** 
    - App Ex 4 
    - Reading (you'll receive an email with the link after class)


# Prepping the data

## Load packages + Paris Paintings data

Load packages:
```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(stringr)
```

Load data:
```{r echo=FALSE}
pp <- read.csv("../../Data/paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

# Recoding variables

## Fixing price

- We did this already!

- Overwrite existing variable `price`:
```{r}
pp <- pp %>%
  mutate(price = as.numeric(str_replace(price, ",", "")))
```

## Shapes of paintings

```{r}
pp %>%
  group_by(Shape) %>%
  summarise(count = n()) %>%
  arrange()
```

## Recode scheme {.smaller}

| original   | new         | change      |
|:-----------|:------------|:------------|
|            | `NA`        | yes         |
| `miniature`| `miniature` | no          |
|   `octagon`| `octagon`   | no          |
|   `octogon`| `octogon`   | yes         |
|      `oval`| `oval`      | no          |
|     `ovale`| `oval`      | yes         |
|  `squ_rect`| `squ_rect`  | no          |
|     `ronde`| `round`     | yes         |
|     `round`| `round`     | no          |

## `ifelse()` function

**Usage**

`ifelse(test, yes, no)`

**Arguments**

`test` - an object which can be coerced to logical mode. <br>
`yes` - return values for true elements of test. <br>
`no` - return values for false elements of test.

- You can nest many `ifelse` statetements.

- See `?ifelse` for more details (just like any other function).

## Recoding shape of painting {.smaller}

Create new variable `shape_recode`:
```{r}
pp <- pp %>%
  mutate(shape_recode = ifelse(Shape == "", NA,
                               ifelse(Shape == "ovale", "oval",
                                      ifelse(Shape == "ronde", "round",
                                             ifelse(Shape == "octogon", "octagon", Shape)))))
```

## Check before you move on!

```{r}
pp %>%
  group_by(shape_recode, Shape) %>%
  summarise(count = n()) %>%
  arrange(shape_recode)
```

## Converting numerical variables to categorical {.smaller}

- If `nfigures` == 0 $\rightarrow$ `fig_mention` = `no figures`
- If `nfigures` >= 1 $\rightarrow$ `fig_mention` = `some figures`

```{r}
# recode
pp <- pp %>%
  mutate(fig_mention = ifelse(nfigures == 0, "no figures", "some figures"))
  
# check
pp %>%
  group_by(fig_mention, nfigures) %>%
  summarise(n())
```

## Recoding, kicked up a notch... {.smaller}

Let's tackle the `mat` variable:

```{r}
pp %>% 
  group_by(mat) %>% 
  summarise(n())
```

## {.smaller}


mat | explanation           | new categories  | mat  | explanation           | new categories
----|:----------------------|:----------------|:-----|:----------------------|:---------------
`a` | silver                | `metal`         | `h`  | oil technique         | uncertain
`al`| alabaster             | `stone`         | `m`  | marble                | stone
`ar`| slate                 | `stone`         | `mi` | miniature technique   | uncertain
`b` | wood                  | `wood`          | `o`  | other                 | other
`bc`| wood and copper       | `metal`         | `p`  | paper                 | paper
`br`| bronze frames         | `uncertain`     | `pa` | pastel                | uncertain    
`bt`| canvas on wood        | `canvas`        | `t`  | canvas                | canvas      
`c` | copper                | `metal`         | `ta` | canvas?               | uncertain 
`ca`| cardboard             | `paper`         | `v`  | glass                 | other   
`co`| cloth                 | `canvas`        | `n/a`| NA                    | NA      
`e` | wax                   | `other`         |      | NA                    | NA    
`g` | grissaille technique  | `uncertain`

## Making use of the `%in%` operator {.smaller}

Create new variable `mat_recode`:
```{r}
pp <- pp %>%
  mutate(mat_recode = ifelse(mat %in% c("a", "bc", "c"), "metal",
                             ifelse(mat %in% c("al", "ar", "m"), "stone",
                                    ifelse(mat %in% c("co", "bt", "t"), "canvas",
                                           ifelse(mat %in% c("p", "ca"), "paper",
                                                  ifelse(mat %in% c("b"), "wood",
                                                         ifelse(mat %in% c("o", "e", "v"), "other",
                                                                ifelse(mat %in% c("n/a", ""), NA,
                                                                       "uncertain"))))))))
```

## Check before you move on!

```{r}
pp %>%
  group_by(mat_recode) %>%
  summarise(count = n()) %>%
  arrange(desc(count))
```


# Formalizing the linear model

## The linear model with a single predictor

>- We're interested in the $\beta_0$ (population parameter for the intercept)
and the $\beta_1$ (population parameter for the slope) in the 
following model:
\[ \hat{y} = \beta_0 + \beta_1~x \]

>- Tough luck, you can't have them...

>- So we use the sample statistics to estimate them:
\[ \hat{y} = b_0 + b_1~x \]

## The linear model with multiple predictors

>- Population model:
\[ \hat{y} = \beta_0 + \beta_1~x_1 + \beta_2~x_2 + \cdots + \beta_k~x_k \]

>- Sample model that we use to estimate the population model:
\[ \hat{y} = b_0 + b_1~x_1 + b_2~x_2 + \cdots + b_k~x_k \]

## Uncertainty around estimates

- Any estimate comes with some uncertainty around it.

- Later in the course we'll discuss how to estimate the uncertainty around
an estimate, such as the slope, and the conditions required for quantifying
uncertainty around estimates using various methods.


# Transformations

## Price vs. surface {.smaller}

<div class="question">
Describe the relationship between price and width of painting.
</div>
```{r fig.width=5,fig.height=3.5}
ggplot(data = pp, aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5)
```

## Let's focus on paintings with `Width_in < 100`

```{r}
pp_wt_lt_100 <- pp %>% 
  filter(Width_in < 100)
```


## Distribution of price {.smaller}

<div class="columns-2">
```{r fig.width=5, fig.height=4, message=FALSE}
ggplot(data = pp_wt_lt_100, aes(x = price)) +
  geom_histogram()
ggplot(data = pp_wt_lt_100, aes(x = log(price))) +
  geom_histogram()
```
</div>

## Price vs. surface {.smaller}

<div class="question">
Which plot shows a more linear relationship?
</div>

<div class="columns-2">
```{r fig.width=4.5, fig.height=3.5, message=FALSE}
ggplot(data = pp_wt_lt_100, 
       aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5)

ggplot(data = pp_wt_lt_100, 
       aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5)
```
</div>

## Transforming the data

>- We saw that `price` has a right-skewed distribution, and the relationship between price and width of 
painting is non-linear.

>- In these situations a transformation applied to the response variable may be useful.

>- In order to decide which transformation to use, we should examine the distribution of the response 
variable.

>- The extremely right skewed distribution suggests that a log transformation may 
be useful.
    - log = natural log ($ln$)

## Logged price vs. surface {.smaller}

<div class="question">
How do we interpret the slope of this model?
</div>

```{r fig.width=7, fig.height=4}
ggplot(data = pp_wt_lt_100, aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = "lm")
```

## Interpreting models with log transformation {.smaller}

```{r}
m_lprice_wt <- lm(log(price) ~ Width_in, data = pp_wt_lt_100)
m_lprice_wt
```

## Linear model with log transformation {.build}

\[ \widehat{log(price)} = 4.67 + 0.02 Width\_in \]

- For each additional inch the painting is wider, the log price of the
painting is expected to be higher, on average, by 0.02 livres.

- which is not a very useful statement...

## Working with logs {.build}

- Subtraction and logs: $log(a) − log(b) = log(a / b)$

- Natural logarithm: $e^{log(x)} = x$

- We can these identities to "undo" the log transformation

## Interpreting models with log transformation {.build}

The slope coefficient for the log transformed model is 0.02, meaning the log 
price difference between paintings whose widths are one inch apart is predicted 
to be 0.02 log livres.

\[log(\text{price for width x+1}) - log(\text{price for width x}) = 0.02 \]

\[log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right) = 0.02 \]

\[e^{log\left(\frac{\text{price for width x+1}}{\text{price for width x}}\right)} = e^{0.02} \]

\[\frac{\text{price for width x+1}}{\text{price for width x}} \approx 1.02\]

For each additional inch the painting is wider, the price of the
painting is expected to be higher, on average, by a factor of 1.02.

## Shortcuts in R

```{r}
m_lprice_wt$coefficients
exp(m_lprice_wt$coefficients)
```


## Recap {.build}

- Non-constant variance is one of the most common model violations, however it 
is usually fixable by transforming the response (y) variable.

- The most common transformation when the response variable is right skewed is 
the log transform: $log(y)$, especially useful when the response variable is 
(extremely) right skewed.

- This transformation is also useful for variance stabilization.

## Recap (cont.) {.build}

- When using a log transformation on the response variable the interpretation of 
the slope changes: "For each unit increase in x, y is expected on average to be 
higher/lower by a factor of $e^{b_1}$."

- Another useful transformation is the square root: $\sqrt{y}$, especially 
useful when the response variable is counts.

- These transformations may also be useful when the relationship is non-linear, 
but in those cases a polynomial regression may also be needed (this is beyond 
the scope of this course, but you’re welcomed to try it for your final project, 
and I’d be happy to provide further guidance).

## Aside: when $y = 0$

In some cases the value of the response variable might be 0, and

```{r}
log(0)
```

The trick is to add a very small number to the value of the response variable 
for these cases so that the $log$ function can still be applied:

```{r}
log(0 + 0.00001)
```


# Application Exercise

## Modeling log transformed prices of Paris Paintings

See course website for details on the application exercise.