---
title: "Sta112FS <br> 6. Data types"
author: "Dr. Çetinkaya-Rundel"
date: "September 15, 2015"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
    css: ../lec.css
---

# Today's agenda

## Today's agenda

- Introduction of teams

- Brief disussion of data types

- Introduction of the Paris Paintings dataset

- Application Exercise: explore Paris Paintings

- **Due Thursday:** Finish App Ex + add info to your GitHub profile

# Data structures and dimensionality

## {.flexbox .vcenter}

----------------------------------------------------
  Dimensions        Homogeneous      Heterogeneous
--------------    ---------------  -----------------
  1*d*              Vector           List
                  (atomic vector)  (generic vector)

  2*d*              Matrix           Data Frame

  n*d*              Array            ---
----------------------------------------------------

# Vectors

## Vector types {.smaller}

R has six basic atomic vector types, but for now we'll only focus on the first four:

* **logical**

* **double**

* **integer**

* **character**

* complex

* raw

## Vector types - examples {.smaller}

**logical** - boolean values `TRUE` and `FALSE`

```{r}
typeof(TRUE)
```

**character** - character strings

```{r}
typeof("hello")
typeof('world')
```

## Vector types - examples {.smaller}

**double** - floating point numerical values (default numerical type)

```{r}
typeof(1.335)
typeof(7)
```

**integer** - integer numerical values (indicated with an `L`)

```{r}
typeof(7L)
typeof(1:3)
```

## Concatenation

Vectors can be constructed using the `c()` function.

```{r}
c(1, 2, 3)
c("Hello", "World!")
c(1, c(2, c(3)))
```


## Coercion

R is a dynamically typed language -- it will happily convert between the various types 
without complaint.

```{r}
c(1, "Hello")
c(FALSE, 3L)
c(1.2, 3L)
```

## Missing Values

R uses `NA` to represent missing values in its data structures.

```{r}
typeof(NA)
```

## Other Special Values {.smaller}

`NaN` - Not a number

`Inf` - Positive infinity

`-Inf` - Negative infinity

<br/>

<div class="columns-2">
```{r}
pi / 0
0 / 0
1/0 + 1/0
1/0 - 1/0
NaN / NA
NaN * NA
```
</div>

## Quick activity: 5 minutes, in teams

What is the type of the following vectors? Explain why they have that type.

* `c(1, NA+1L, "C")`
* `c(1L / 0, NA)`
* `c(1:3, 5)`
* `c(3L, NaN+1L)`
* `c(NA, TRUE)`

# Lists

## Lists {.smaller}

Lists are _generic vectors_: 1d and can contain any combination of R objects.

```{r}
mylist = list("A", 1:4, c(TRUE, FALSE), (1:4)/2)
mylist
```

```{r}
str(mylist)
```

## Named lists {.smaller}

Because of their more complex structure we often want to name the elements of a list (we 
can also do this with vectors). This can make reading and accessing the list more 
straight forward.

```{r}
myotherlist <- list(A = "hello", B = 1:4, "knock knock" = "who's there?")
str(myotherlist)
names(myotherlist)
myotherlist$B
```

## Vectors vs. lists - [ vs. [[ {.smaller}

```{r, error=TRUE}
x <- c(8,4,7)
x[1]
x[[1]]
y <- list(8,4,7)
y[2]
y[[2]]
```

# Data Frames

## Data Frames

A data frame is the most commonly used data structure in R, they are just a list of equal length vectors (usually atomic, but you can use generic as well). Each vector is treated as a column and elements of the vectors as rows.

Most often a data frame will be constructed by reading in from a file, but we can also create them from scratch.

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

## {.smaller}

```{r}
attributes(df)
class(df$x)
class(df$y)
```

## Strings (Characters) vs Factors

By default R will often attempt to convert character vectors into factors when they are 
included in a data frame. Sometimes this is useful, sometimes it isn't -- either way it 
is important to know what type/class you are working with. This behavior can be changed 
using the `stringsAsFactors` argument.

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"), stringsAsFactors = FALSE)
str(df)
```

## Growing data frames {.smaller}

We can add rows or columns to a data frame using `rbind` and `cbind` respectively.

```{r}
df <- data.frame(x = 1:3, y = c("a","b","c"))
cbind(df, z = TRUE)
rbind(df, c(4,"b"))
```


## Combining data frames {.smaller}

```{r}
df1 <- data.frame(x = 1:3, y = c("a", "b", "c"))
df2 <- data.frame(m = 3:1, n = c(TRUE, TRUE, FALSE))
df3 <- cbind(df1, df2)
df3
str(df3)
```

## `tbl_df()` in `dplyr` {.smaller}

```{r message = FALSE}
library(dplyr)
```

```{r}
df1 <- data.frame(x = 1:3, 
                  y = c("a", "b", "c"))
tbl_df(df1)
```

## `tbl_df()` in `dplyr` (cont.) {.smaller}

```{r}
df2 <- data.frame(m = 3:1, 
                  n = c(TRUE, TRUE, FALSE)) %>%
  tbl_df()
df2
```

# Summary of data structures

## Recap

- Always best to think of data as part of a data.frame
    + This plays nicely with `dplyr` and `ggplot2` as well
    + Rows are observations, columns are variables
    
- Be careful about data types / classes
    + Sometimes `R` makes silly assumptions about your data class 
        + `stringsAsFactors = FALSE` helps, but that's not the whole story
        + 0/1 variable is most likely a `factor`
    + If a plot/output is not behaving the way you expect, first
    investigate the data class
    + If you are absolutely sure of a data class, over-write it in your
    data frame so that you don't need to keep having to keep track of it
        + `mutate` the variable with the correct class

## Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)

# Paris Paintings

## Paintings dataset

Click [here](https://stat.duke.edu/courses/Fall15/sta112.01/post/slides/STAT112_ParisPaintings.pdf) for Sandra's slides

## Data sharing

This dataset is made available for class use only. Do not post the raw dataset
anywhere else. You can share your results/findings/plots etc. but not the dataset.

## Accessing the data

- Codebook: https://stat.duke.edu/courses/Fall15/sta112.01/data/paris_paintings.html

- Go to the [Resources on Sakai](https://sakai.duke.edu/portal/site/a91015e2-86df-49b6-9c16-e83a9ca5da66/page/fb4fe022-ccfe-4806-a4ff-4d73fdb1bf08) and download `paris_paintings.csv`

- Upload this file to RStudio Server

- Load using the following (make sure data file is in the correct working
directory):

```{r eval=FALSE}
pp <- read.csv("paris_paintings.csv", stringsAsFactors = FALSE) %>%
  tbl_df()
```

# Application exercise

## App Ex 2

See course website