---
title: "Course evaluations"
output:
  html_document:
    highlight: pygments
    theme: flatly
    css: ../hw.css
  
references:
- id: Hamermesh2005
  title: Beauty in the Classroom - Instructors’ Pulchritude and Putative Pedagogical Productivity
  author:
  - family: Hamermesh
    given: Daniel S.
  - family: Parker
    given: Amy
  volume: 24
  URL: 'http://www.sciencedirect.com/science/article/pii/S0272775704001165'
  DOI: 10.1016/j.econedurev.2004.07.013
  publisher: Economics of Education Review
  ISSN: 0272-7757
  issue: 4
  page: 369-376
  type: article-journal
  issued:
    year: 2005
    month: 8
- id: Gelman2007
  title: Data Analysis Using Regression and Multilevel/Hierarchical Models
  author:
  - family: Gelman
    given: Andrew
  - family: Hill
    given: Jennifer
  publisher: Cambridge University Press
  city:
  type: book
  issued:
    year: 2007
  edition: 1
  ISBN: 9780521686891
---


```{r opts, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Instructions

1. Start with your repo for this assignment
    - located at [organization](https://github.com/Sta112FS-Fall2015) for this class
    - name contains HW 2 and your name
    
Clone this repo in your local directory on `gort`. (Remember, the 
address is http://gort.stat.duke.edu:8787/.)

2. Edit the README.md to include some relevant information about the
repository, commit, and push. (This is just to check everything
is working fine, and you know what you're doing.)

3. Open a new R Markdown file, name it the same name as your repository,
and save it.

4. Include answers to all exercises in your R Markdown file. Your
answers should always include any summary and/or plot you use to answer
that particular question.

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" [@Hamermesh2005] 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. 

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.


## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is a slightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
[@Gelman2007].) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and 
professors.

Let's first load the packages we'll use for this exercise. Note that you might
not have some of them installed. If that's the case, use `install.packages` to
install them first.
```{r load-packages, message=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(GGally)
```

And let's also load the data:
```{r load-data}
load(url("http://www.openintro.org/stat/data/evals.RData"))
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not? Include any summary statistics and visualizations
    you use in your response.

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualizations.

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg}
ggplot(data = evals, aes(x = bty_avg, y = score)) +
  geom_point()
```

Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

4.  Replot the scatterplot, but this time use `geom_point(position = "jitter")`.
    What does "jitter" mean? What was misleading about the initial scatterplot?

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the regression line to your
    plot. Write out the equation for the linear model and
    interpret the slope. Does there appear to be a practically significant
    relationship between professor score and average beauty rating?

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
ggplot(data = evals, aes(x = bty_avg, y = bty_f1lower)) +
  geom_point(position = "jitter")
evals %>% 
  summarise(cor(bty_avg, bty_f1lower))
```

As expected the relationship is quite strong. Note that the correlation coefficient
measures the linear association between two variables, and ranges between -1 and 1, 
-1 indicating a perfect negative relationship and 1 indicating a perfect positive
relationship. It is nor surprising that the correlation between these two variables 
is quite strong - after all, the average score is calculated using the individual 
scores. We can actually take a look at the relationships between all beauty 
variables (variables that have the character string `bty` in them) using the 
following command:

```{r bty-rels}
bty_cols <- str_detect(names(evals), "bty_")
ggpairs(evals, columns = which(bty_cols == TRUE))
```

6.  Describe how the `str_detect` function works and how we use information
    resulting from that function to make the pairwise plot.

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a a good predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r model-bty_avg_gender-main}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
m_bty_gen
```

7.  How do the AIC and the adjusted $R^2$ of this model and the previous one
    compare? Has the addition of `gender` to the model changed the parameter
    estimate (slope) for `bty_avg`?

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

8.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?
    
The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)
    
9.  Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

We can go a step further and include an interaction variable between `bty_avg`
and `gender`:

```{r model-bty_avg_gender-int}
m_bty_gen_int <- lm(score ~ bty_avg * gender, data = evals)
m_bty_gen_int
```

We can also visualize this model with the following:

```{r twoLines}
ggplot(data = evals, aes(x = bty_avg, y = score, color = gender)) +
  geom_point() +
  stat_smooth(method = "lm")
```

10. How does the relationship between beauty and evaluation score
    vary between male and female professors?

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to be the worst predictor of evaluation scores?
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

Let's run the model...

```{r m_full, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

For simplicity we'll work with a model that does not have interaction effects.

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

13. Interpret the coefficient associated with the ethnicity variable.

14. Drop the variable that results in the highest gain of adjusted R-squared.
    Did the coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

15. Using backward-selection and either AIC or adjusted R-squared as the selection 
    criterion, determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

16. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.

17. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?

## License 
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was written by 
Mine &Ccedil;etinkaya-Rundel and Andrew Bray.

## References