---
title: "Sta112FS <br> 11. Model selection, prediction, and model validation"
author: "Dr. Çetinkaya-Rundel"
date: "October 1, 2015"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
    css: ../lec.css
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 90)
#knitr::opts_chunk$set(cache = TRUE)
```

# Today's agenda

## Today's agenda

- Answer any project related questions

- Recap App Ex 5
 
- Model selection, prediction, and model validation
    - Model selection with `step` and AIC
    - Prediction
    - Model validation

- Before leaving class: make appoinment to meet with me as a 
team during my Monday OH

- **Due Tuesday:** Work on project!


# Prediction

## Setup

Load packages:
```{r load-packages, message=FALSE}
library(ggplot2)
library(dplyr)
library(nycflights13) # may need to install first
```

Load data from `nycflights13` package:
```{r load-data}
data(flights)
```

## Take a look at the `flights` data {.smaller}

```{r str-data}
str(flights)
```

## Goal: Predict airline delays - `arr_delay` {.smaller}

<div class="question">
Describe the distribution of arrival delays. What does `arr_delay = 0` mean?
</div>

```{r message=FALSE, fig.height=3.5, fig.width=5.5}
ggplot(data = flights, aes(x = arr_delay)) +
  geom_histogram() +
  xlab("arrival delay, mins")
```

## Only consider flights that are actually delayed

Let's say "actually delayed" means delayed more than 1 minute:
```{r}
delayed_flights <- flights %>%
  filter(arr_delay > 1)
```

There are `r nrow(delayed_flights)` such flights in the sample:
```{r}
nrow(delayed_flights)
```

## Log transformation? {.smaller}

<div class="columns-2">
```{r fig.width=5, fig.height=3,message=FALSE}
ggplot(data = delayed_flights, 
       aes(x = arr_delay)) +
  geom_histogram() +
  xlab("arrival delay, mins")

ggplot(data = delayed_flights, 
       aes(x = log(arr_delay))) +
  geom_histogram() +
  xlab("logged arrival delay, log(mins)")
```
</div>

## Get a feeling for the data

On average flights are delayed 42 minutes. The median is 22 minutes, and the middle
50% of flights are delayed 9 to 53 minutes. The flight that is the most delayed has
arrived 21.2 hours late at its destination.

```{r}
delayed_flights %>%
  summarise(mean_mins = mean(arr_delay), median_mins = median(arr_delay), 
            Q1_mins = quantile(arr_delay, 0.25), Q3_mins = quantile(arr_delay, 0.75),
            max_hrs = max(arr_delay)/60)
```

## Candidate variables {.smaller}

- Variables to exclude: `year` since they're all 2013, `dep_delay` since that would be 
cheating, `tailnum` and `flight` since they are not very informative, `dep_time` since it 
is redundent with below, and `arr_time` and `air_time` since we wouldn't know these before
taking the flight

- Create a new variable from `hour` and `minute`, $(hour \times 60 + minute)$, so that
a linear increase of by a minute in departure time can be easily interpreted
```{r create-time_in_mins, fig.height=3,fig.width=5,message=FALSE}
delayed_flights <- delayed_flights %>% mutate(time_in_mins = hour * 60 + minute)
ggplot(data = delayed_flights, aes(x = time_in_mins)) + geom_histogram()
```


## Fit full model

```{r}
full_model <- lm(log(arr_delay) ~ month + day + carrier + origin + dest + 
                   distance + time_in_mins, data = delayed_flights)
summary(full_model)$r.squared
```

## Akaike Information Criterion

\[ AIC = -2log(L) + 2k \]

- $L$: likelihood	of the	model, i.e.	likelihood of seeing these data	given	the	model	
parameters	
- Applies	a	penalty	for	number of parameters in the	model, $k$	
- Used for model selection, lower the better
    - Value is not informative on its own
- In R to get the AIC, use `AIC(model)`

```{r aic-full-model}
AIC(full_model)
```

## Model selection -- a little faster

```{r results="hide"}
final_model <- step(full_model, direction = "backward")
```



```{r}
AIC(final_model)
```

## Parsimony

<div class="question">
Take a look at the variables in the full and the final model. Can you guess
why some of them may have been dropped? Remember: We like parsimonous models.
</div>

```{r}
full_model$call
final_model$call
```


# Prediction

## New observation

Suppose I'm taking a flight from JFK to RDU, in October, with Jet Blue (`B6`), and I'm taking the flight at 4pm ($16 * 60 + 0 = 960$).

```{r}
mine_flight <- data.frame(month = 10, carrier = "B6", origin = "JFK", dest = "RDU", distance = 427, time_in_mins = 960)
```

## Prediction

```{r}
exp(predict(final_model, newdata = mine_flight))
```

# Model validation

## Model validation

- One commonly used model validation technique is to partition your data into training
and testing set

- That is, fit the model on the training data

- And test it on the testing data

- Evaluate model performance using $RMSE$, root-mean squared error

\[ \sqrt{\frac{\sum_{i = 1}^n (y_i - \hat{y}_i)^2}{n}} \]

<div class="question">
Do you think we should prefer low or high RMSE?
</div>

## Random sampling and reproducibility

Gotta set a seed!
```{r}
set.seed(10012015)
```

- Don't reuse your seeds

- Use different seeds from each other

- Need inspiration? https://www.random.org/


## 80% training, 20% testing

Find index numbers for the observations to be included in the trainining dataset:
```{r}
n <- nrow(delayed_flights)
training_index <- sample(1:n, size = n * 0.80, replace = FALSE)
```

<br>

<div class="columns-2">
Create training dataset:
```{r}
training_data <- delayed_flights %>%
  slice(training_index)
nrow(training_data)
```
<br>
Create testing dataset:
```{r}
testing_data <- delayed_flights %>%
  slice(-training_index)
nrow(testing_data)
```
</div>

## Your turn! {.smaller}

```{r eval=FALSE}
library(ggplot2); library(dplyr); library(nycflights13)
set.seed(123) #insert some value for the seed

# delayed flights
delayed_flights <- flights %>%
  filter(arr_delay > 1)
nrow(delayed_flights)
dim(delayed_flights)
names(delayed_flights)

# time_in_mins
delayed_flights <- delayed_flights %>% 
  mutate(time_in_mins = hour * 60 + minute)
dim(delayed_flights)

# training indices
n <- nrow(delayed_flights)
training_index <- sample(1:n, size = n * 0.80, replace = FALSE)
length(training_index)

# training data
training_data <- delayed_flights %>%
  slice(training_index)
nrow(training_data)

# testing data
testing_data <- delayed_flights %>%
  slice(-training_index)
nrow(testing_data)
```

## Step 1: Fit the full model on your training data

```{r fitting-model}
full <- lm(log(arr_delay) ~ month + day + carrier + origin + dest + 
                   distance + time_in_mins, data = training_data)
full
```

<div class="question">
Compare the output of your full model with neighbors. Are the parameter
estimates the same? Different? Very different?
</div>

## Step 2: Model selection

```{r}
final <- step(full, direction = "backward")
```

<div class="question">
Compare your final model with mine. Do we have the same variables chosen?
</div>
```{r}
final$call
```

## Step 3: Prediction and RMSE

Calculate predicted delay time, and don't forget to exponentiate!
```{r}
log_y_hat <- predict(final, newdata = testing_data)
y_hat <- exp(log_y_hat)
```

Calculate RMSE
```{r}
rmse <- sqrt(sum((y_hat - testing_data$arr_delay)^2) / nrow(testing_data))
rmse
```

<div class="question">
What is your RMSE? How does it compare to others'?
</div>

