---
title: "Sta112FS <br> 4. Tidy data & Intro to data visualization"
author: "Dr. Çetinkaya-Rundel"
date: "September 3, 2015"
output:
  html_document:
    highlight: pygments
    widescreen: yes
    css: ../lec.css
---

```{r echo=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
gapminder <- read.csv("https://stat.duke.edu/~mc301/data/gapminder.csv")
```

# From last time...

## A note on piping and layering

- The `%>%` operator in `dplyr` functions is called
the pipe operator. This means you "pipe" the output of the previous
line of code as the first input of the next line of code.

- The `+` operator in `ggplot2` functions is used for "layering". This
means you create the plot in layers, separated by `+`.

# Today's agenda

## Today's agenda

- Highlights from the "Tidy data" paper

- Visualizing data with `ggplot2`
    - **Goal:** Learn syntax and various visualizations for different
    data types

# Tidy data

## Tidy data {.smaller}

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

<div class="question">
Which of the following is a data set and which is a summary
table?
</div>

<div class="columns-2">

```{r echo=FALSE}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(mean_lifeexp = mean(lifeExp)) %>%
  tbl_df()
```

<br><br><br><br>

```{r echo=FALSE}
gapminder %>%
  filter(year == 2007) %>%
  select(country, continent, lifeExp) %>%
  tbl_df()
```

</div>

# Data visualization

## `ggplot2`

- To use `ggplot2` functions, first load the package
```{r}
library(ggplot2)
```

- In `ggplot2` the structure of the code for plots can often be 
summarized as

```{r eval = FALSE}
ggplot + 
  geom_xxx
```

or, more precisely
```{r eval = FALSE}
ggplot(data = [dataset], aes(x = [x-variable], [y-variable])) +
   geom_xxx() +
   other options
```

- Geoms, short for geometric objects, describe the type of plot you will produce

## Context: Fuel economy {.build}

**data:** `mpg` - fuel economy data from 1999 and 2008 for 38 popular 
models of cars

- Run the following **in the Console**:
```{r eval = FALSE}
# see the help file
?mpg
```

```{r eval = FALSE}
# view data
View(mpg)
```

<div class="question">
How many rows and columns does this dataset have? What does each row represent? What does each column represent?
</div>


<div class="question">
Make a prediction: What relationship do you expect to see between engine size (displ) and mileage (hwy)?
</div>

# Scatterplots

## Displacement vs. highway mpg (`geom_point()`)

<div class="question">
How would you describe this relationship? What other variables would help us understand data points that don't follow the overall trend?
</div>

```{r fig.width = 6, fig.height=3.5}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Additional variables

Can display additional variables with

- aesthetics (like shape, colour, size), or

- faceting (small multiples displaying different subsets)

# Aesthetics

## Aesthetics options

Visual characteristics of plotting characters that can be **mapped to data** are

- `color`

- `size`

- `shape`

- `alpha` (transparency)

## Displacement vs. highway mpg + class

```{r fig.width = 6, fig.height=3.5}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Aesthetics summary

aesthetics    | discrete     | continuous
------------- | ------------ | ------------
color         | rainbow of colors | gradient
size          | discrete steps    | linear mapping between radius and value
shape         | different shape for each | shouldn't (and doesn't) work

# Faceting

## Faceting options

- Smaller plots that display different subsets of the data

- Useful for exploring conditional relationships and large data

## Displacement vs. highway mpg by cylinders

```{r fig.height=4, fig.width=8}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl) +
  geom_point()
```

## Dive further...

<div class="question">
In the next few slides describe what each plot displays. Think about
how the code relates to the output.
</div>

## 

```{r eval=TRUE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .) +
  geom_point()
```

## 

```{r eval=TRUE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl) +
  geom_point()
```

##

```{r eval=TRUE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  facet_wrap(~ class) +
  geom_point()  
```

## Facet summary

- `facet_grid()`: 2d grid, rows ~ cols, . for no split

- `facet_wrap()`: 1d ribbon wrapped into 2d

# Other geoms

## Displacement vs. highway mpg, take 2

<div class="question">
How are these plots similar? How are they different?
</div>

```{r echo = FALSE, message=FALSE, warning=FALSE}
p1 = qplot(displ, hwy, data = mpg)
p2 = qplot(displ, hwy, data = mpg, geom = "smooth")
suppressMessages(suppressWarnings(require(gridExtra)))
grid.arrange(p1, p2, ncol = 2)
```

## `geom_smooth`

To plot a smooth curve, use `geom_smooth()`

```{r message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_smooth()
```

# Exploratory data analysis (EDA)

## Number of variables involved

* Univariate data analysis - distribution of single variable

* Bivariate data analysis - relationship between two variables

* Multivariate data analysis - relationship between many variables at once, usually focusing on the relationship between two while conditioning for others

## Types of variables

- **Numerical variables** can be classified as **continuous** or **discrete** based on whether or not the variable can take on an infinite number of values or only non-negative whole numbers, respectively. 

- If the variable is **categorical**, we can determine if it is **ordinal** based on whether or not the levels have a natural ordering.

## Describing shapes of numerical distributions

* shape:
    * skewness: right-skewed, left-skewed, symmetric (skew is to the side of the longer tail)
    * modality: unimodal, bimodal, multimodal, uniform
* center: mean (`mean`), median (`median`), mode (not always useful)
* spead: range (`range`), standard deviation (`sd`), inter-quartile range (`IQR`)
* unusal observations

[Put these, and more, to use in HW 1]

# Homework

## Homework 1

- Can collaborate with others, but must submit own work

- Submission on GitHub (follow instructions on HW)

- **Due:** By class on Thursday (Sep 11)
    - Will get a chance to work on it in class on Tuesday, but make
    sure to at least have read over it before then
    - Prof. C-R hours on Monday 3:30 - 5:30pm
    - Ask questions on Piazza