---
title: "Sta112FS <br> 3. R/RStudio/GitHub demo"
author: "Dr. Çetinkaya-Rundel"
date: "September 1, 2015"
output:
  html_document:
    highlight: pygments
    widescreen: yes
    css: ../lec.css
---

# Housekeeping

## Announcements

- HW 1 is due next Tuesday

# From last time...

## Task

Visualize relationship between life expectancy and GDP per capita in
2007 in countries. Also create a visualization 

## Step 0: Load necessary packages

- In the following exercises we'll use `dplyr` (for data wrangling) and 
`ggplot2` (for visualization) packages.

- Make sure the packages are installed.

- Load these packages in your markdown file:

```{r message=FALSE}
library(dplyr)
library(ggplot2)
```

## Step 1: Load data

```{r}
gapminder <- read.csv("https://stat.duke.edu/~mc301/data/gapminder.csv")
```

## Step 2: Subset data

- Start with the `gapminder` dataset

- Filter for cases (rows) where year is equal to 2007

- Save this new subsetted dataset as `gap07`

```{r}
gap07 <- gapminder %>%
  filter(year == 2007)
```


## Step 2: Explore and visualize

**Task:** Visualize the relationship between `gdpPercap` and `lifeExp`.

```{r, fig.height=4, fig.width=7}
qplot(x = gdpPercap, y = lifeExp, data = gap07)
```

## Step 3: Dig deeper

**Task:** Color the points by continent.

```{r, fig.height=4, fig.width=7}
qplot(x = gdpPercap, y = lifeExp, color = continent, data = gap07)
```

## Step 3: Commit and push all changes

- Stage

- Commit (with a message)

- Push

## A note on piping and layering

- The `%>%` operator in `dplyr` functions is called
the pipe operator. This means you "pipe" the output of the previous
line of code as the first input of the next line of code.

- The `+` operator in `ggplot2` functions is used for "layering". This
means you create the plot in layers, separated by `+`.

# What's next?

## Update your analysis

What if you wanted to now change your analysis 

- to subset for 1952

- plot life expectancy (`lifeExp`) vs. population (`pop`)

- and size the points by GPD (`gpdPercap`)
    - hint: add argument `size = gpdPercap` to your plotting code

Once you're done, commit and push all your changes with a meaningful
message.